---
title: "ProteinDynamics_DimReduction"
author: "Arshan Munif (2262291)"
output:
  word_document: default
  html_document: default
date: "2023-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About the R Markdown

This R Markdown file is designed to load and analyse protein trajectory data and conduct dimensionality reduction using PCA. To begin the user will read .pdb and .xtc files in this markdown

```{r}
# Get working directory
# The .pdb and .xtc files need to be in this directory. If not, R will not be able to locate and read the files and produce errors.
getwd()
```

### Load Libraries

```{r}
# Load necessary libraries
library(bio3d)
library(rioxdr)
library(ggplot2)
library(ggfortify)
library(MASS)
```


### 1. Read trajectory and topology file, Select and extract C-alpha atoms, Reshape the data, Compute RMSD and PCA results

```{r}
# Source the functions
source('trajectory_function.R')
source('filter_CAlpha.R')
source('Reshape_Data.R')
source('extract_coords.R')
source('RMSD.R')
source('PCA_func.R')

# Use the function to read trajectory(xtc) and topology (pdb) files
trj_obj <- read_xtc("MD05_ADK_data/MD05_ADK_protein_fit_small.xtc", "MD05_ADK_data/MD05_ADK_protein.pdb")

# Select c-alpha atoms
inds <- select_calpha(trj_obj$pdb)

# Extract c-alpha atom coordinates
ca_coords <- extract_calpha_coords(trj_obj$trj, inds)


# Reshape the data
#reshaped_coords <- reshape_data(ca_coords, inds)
reshaped_coords <- ca_reshape(ca_coords)


# check the dimensions to ensure data prep and transformation were performed properly
dim(ca_coords)
dim(reshaped_coords)

# Compute RMSD
rmsd_values <- compute_rmsd(reshaped_coords)
mean(rmsd_values)
hist(rmsd_values)


# Perform PCA using the PCA function
pca_result <- perform_pca(reshaped_coords)
plot(pca_result, col= 'orange', xlab = 'Principal Components', main = 'PCA Results Plot') # this plot will show the principal components in descending order of variance

```


### 2. Extract and plot principal components, compute explained variance, cumulative PVE, and visualise loadings

```{r}
# Extract principal components scores
scores <- pca_result$x
head(scores[,1:3]) # prints the first few lines for selected PCs
tail(scores[,1:3]) # prints the last few lines for selected PCs
plot(scores, main = 'PC Scores') # plot the scores

# Generate color coded scores plot
rgb.palette <- colorRampPalette(c("blue","green", "orange"), space = 'rgb')
plot(scores[,1], scores[,2], pch = 20, col = rgb.palette(1000), xlab = 'PC1 Scores', ylab = 'PC2 Scores')

# Generate kernel density map
a = kde2d(scores[,1], scores[,2], n = 1000)
image(a)

# calculate proportion of variance explained (PVE) from std values
pca_result_var <- pca_result$sdev^2
cat("Variance for selected PCs:", pca_result_var[1:3], "\n")

pca_result_PVE <- pca_result_var / sum(pca_result_var)
cat("Explained variance (PVE) for selected PCs:", pca_result_PVE[1:3], "\n")


# calculate cumulative value of PVE for increasing number of additional principal components
cumul_var <- cumsum(pca_result_PVE)
cat("Cumulative PVE for selected PCs:", cumul_var[1:3], "\n")


# get and inspect loadings principal components
pca_loadings <- pca_result$rotation
cat("Loadings for the selected PCs:", pca_loadings[1:3], "\n")


# plot the loadings for the PCs
source('Loading_Plot.R') # call loading plot function
selected_components <- c(1, 2, 3)
plot_loadings(pca_loadings, selected_components)
```

### 3. Generate PVE plot and Identify the top principal components representing signa
l
```{r}
# Call the PVE plot function
source('PVE_Plot.R')

# Generate plot with a 80% threshold line to inform feature extraction
selected_components <- 5 # insert the desired number of PCs that you want to be visualised in the scree plot
select_pc <- pve_plot(cumul_var, selected_components)

# Print the number of components that explain 80% of variance
cat('Selected number of components:', select_pc)
```


### 4. Generate Bi-plot

```{r}
library(factoextra)

# call the biplot function
source('biplot_func.R') 

num_scores <- 1000  # Change this to the desired number of scores that needs to be shown
subset_pca_result <- pca_result
subset_pca_result$x <- pca_result$x[1:num_scores, ]

# Call the pca_biplot function
pca_biplot(subset_pca_result)

# visualise the variables
fviz_pca_var(pca_result, select.var = list(contrib = 100))
```

### 5. Project energy values on the PCA space

```{r}
# read the energy table
energy <- read.table("/Users/arshanmunif/Documents/GitHub/MSRDimRed/MD05_small.energy")

```

```{r}
# create a rgb colour coding function
function (n) 
{
    x <- ramp(seq.int(0, 1, length.out = n))
    if (ncol(x) == 4L) 
        rgb(x[, 1L], x[, 2L], x[, 3L], x[, 4L], maxColorValue = 255)
    else rgb(x[, 1L], x[, 2L], x[, 3L], maxColorValue = 255)
}
rgb.palette(1000)
```

```{r}
# Generate a colour vector ordered by energy values
col_vec <- rgb.palette(1000)[order(energy$V1)]

# Define a custom colour palette
rgb.palette <- colorRampPalette(c("blue","green", "orange"),space = 'rgb')

# Generate colour coded energy projection plot
plot(scores, col = col_vec, pch = 16, main = "Energy Projection on PCs")

```

### 6. Project RMSD Values on the PCA Space

```{r}

# Define a custom colour palette
rgb.palette <- colorRampPalette(c("blue", "green", "orange"), space = 'rgb')

# Generate a colour vector ordered by RMSD values
col_vec_2 <- rgb.palette(length(rmsd_values))[order(rmsd_values)]

# project the rmsd on PCA plot
plot(scores, col = col_vec_2, pch = 16, main = "RMSD Values Projected on PCs")


```

### 7. Project atomic distances on the PCA space

```{r}
distance <- read.table("/Users/arshanmunif/Documents/GitHub/MSRDimRed/MD05_ADK_protein_fit_small.G55-P127.xvg", skip=17)
names(distance) <- c("time", "distance")
distance$col <- rgb(0,
                    (
                      (distance$distance - min(distance$distance)) / 
                        (max(distance$distance) - min(distance$distance))),
                     0)

# geneate histogram to understand distribution
hist(distance$distance, 100)
hist(distance$distance, 1000)

# assign colours to distinguish open and closed states of protein
distance$open <- 'red' # red depicts closed protein state
distance[distance$distance > 2.1, 'open'] <- 'blue' # blue depicts open protein state

# check colour assignment by viewing the first 6 rows in the data frame
head(distance)

# generate plot with the distance values projected on the PCA space
plot(scores, col = distance$open, pch = 16, main = "Distances Projected on PCs") 
contour(a, add = T) # add contour overlay to the PCA plot to visualise the minima
```



















